---
title: "STAT 223 Final Project"
author: "Lee Kapp"
date: "5/15/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
lapply(c("readxl","tidyverse", "magrittr", "GGally", "car", "caret", "corrplot", "randomcoloR", "skimr", "kableExtra", "psych", "MVN", "ggrepel", "SimComp", "heplots", "cluster", "NbClust", "factoextra"), library, character.only=T)
options(digits=4)
purps <- randomColor(count = 7, hue = "purple")
greens <- randomColor(count = 6, hue = "green")
reds <- randomColor(count = 2, hue = "red")
blues <- randomColor(count = 3, hue = "blue")
palette <- c(purps, "yellow", "orange", greens, reds, blues)
names(palette) <- c("V", "I", "L", "G", "A", "F", "W", "C", "M", "P", "Y", "T", "S", "N", "Q", "D", "E", "H", "K", "R")
print(purps)
print(greens)
print(reds)
print(blues)
```


### Data Import
```{r data}
data <- read_xls("/Users/leekapp/Documents/Portfolio/STAT223Final/AminoAcidAbundanceVertebrates.xls", sheet = "Combined")
ER <- data %>% filter(location == "ER")
golgi <- data %>% filter(location == "Golgi")
tgnEndo <- data %>% filter(location == "TGNEndo")
lyso <- data %>% filter(location == "lysosomes")
PM <- data %>% filter(location == "PM_total")
```

### Descriptive Stats
```{r description}
erMin <- apply(ER[-c(1,22)], 2, min)
erMean <- apply(ER[-c(1,22)], 2, mean)
erSD <- apply(ER[-c(1,22)], 2, mean)
erMax <- apply(ER[-c(1,22)], 2, max)

ERsumm <- data.frame(rbind(erMin, erMean, erSD, erMax))
rownames(ERsumm) <- c("min", "mean", "std dev", "max")
ERsumm <- round(ERsumm, 4)

golgiMin <- apply(golgi[-c(1,22)], 2, min)
golgiMean <- apply(golgi[-c(1,22)], 2, mean)
golgiSD <- apply(golgi[-c(1,22)], 2, mean)
golgiMax <- apply(golgi[-c(1,22)], 2, max)

golgiSumm <- data.frame(rbind(golgiMin, golgiMean, golgiSD, golgiMax))
rownames(golgiSumm) <- c("min", "mean", "std dev", "max")
golgiSumm <- round(golgiSumm, 4)

lysoMin <- apply(lyso[-c(1,22)], 2, min)
lysoMean <- apply(lyso[-c(1,22)], 2, mean)
lysoSD <- apply(lyso[-c(1,22)], 2, mean)
lysoMax <- apply(lyso[-c(1,22)], 2, max)

lysoSumm <- data.frame(rbind(lysoMin, lysoMean, lysoSD, lysoMax))
rownames(lysoSumm) <- c("min", "mean", "std dev", "max")
lysoSumm <- round(lysoSumm, 4)

pmMin <- apply(PM[-c(1,22)], 2, min)
pmMean <- apply(PM[-c(1,22)], 2, mean)
pmSD <- apply(PM[-c(1,22)], 2, mean)
pmMax <- apply(PM[-c(1,22)], 2, max)

pmSumm <- data.frame(rbind(pmMin, pmMean, pmSD, pmMax))
rownames(pmSumm) <- c("min", "mean", "std dev", "max")
pmSumm <- round(pmSumm, 4)

tgnMin <- apply(tgnEndo[-c(1,22)], 2, min)
tgnMean <- apply(tgnEndo[-c(1,22)], 2, mean)
tgnSD <- apply(tgnEndo[-c(1,22)], 2, mean)
tgnMax <- apply(tgnEndo[-c(1,22)], 2, max)

tgnSumm <- data.frame(rbind(tgnMin, tgnMean, tgnSD, tgnMax))
rownames(tgnSumm) <- c("min", "mean", "std dev", "max")
tgnSumm <- round(tgnSumm, 4)

ERsumm %>% 
  kbl(caption = "Summary Statistics by Amino Acid: Endoplasmic Reticulum") %>% 
  kable_classic(full_width = F, html_font = "Courier", font_size = 10) %>% 
  column_spec(1, bold = T) %>% 
  row_spec(0, bold = T) %>% 
  save_kable("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/ERsumm.pdf")

golgiSumm %>% 
  kbl(caption = "Summary Statistics by Amino Acid: Golgi") %>% 
  kable_classic(full_width = F, html_font = "Courier", font_size = 10) %>% 
  column_spec(1, bold = T) %>% 
  row_spec(0, bold = T) %>% 
  save_kable("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/golgiSumm.pdf")

lysoSumm %>% 
  kbl(caption = "Summary Statistics by Amino Acid: Lysosomes") %>% 
  kable_classic(full_width = F, html_font = "Courier", font_size = 10) %>% 
  column_spec(1, bold = T) %>% 
  row_spec(0, bold = T) %>% 
  save_kable("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/lysoSumm.pdf")

tgnSumm %>% 
  kbl(caption = "Summary Statistics by Amino Acid: TGN:Endosomes") %>% 
  kable_classic(full_width = F, html_font = "Courier", font_size = 10) %>% 
  column_spec(1, bold = T) %>% 
  row_spec(0, bold = T) %>% 
  save_kable("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/tgnSumm.pdf")

pmSumm %>% 
  kbl(caption = "Summary Statistics by Amino Acid: Plasma Membrane") %>% 
  kable_classic(full_width = F, html_font = "Courier", font_size = 10) %>% 
  column_spec(1, bold = T) %>% 
  row_spec(0, bold = T) %>% 
  save_kable("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/pmSumm.pdf")
```



### EDA plots - need a loop to make one plot per amino acid and then facet by amino acid
```{r eda plots}

vars <- colnames(data)[2:21]
for(i in vars) {
  plt <- ggplot(data = data, mapping = aes_string(x = "Position", y = i,  color = "location")) +
    geom_point(size = 2) +
    theme_minimal() +
    labs(title = paste("Plot of ", i), y = paste("Amino Acid ", i)) +
    facet_wrap(~location, nrow = 4) +
    ggsave(paste("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/EDAplot ", i, ".pdf"))
    
    #print(plt)
}
```


### Variances and Correlations 
```{r vars}
#checking variances
ER_S <- var(ER[-c(1,22)])
golgi_S <- var(golgi[-c(1,22)])
lyso_S <- var(lyso[-c(1,22)])
PM_S <- var(PM[-c(1,22)])
tgnEndo_S <- var(tgnEndo[-c(1,22)])

#They're all tiny
print("ER")
print(diag(ER_S))
print("Golgi")
print(diag(golgi_S))
print("Lysosomes")
print(diag(lyso_S))
print("PM")
print(diag(PM_S))
print("TGN:Endo")
print(diag(tgnEndo_S))

#checking correlations
ER_R <- cor(ER[-c(1,22)])
golgi_R <- cor(golgi[-c(1,22)])
lyso_R <- cor(lyso[-c(1,22)])
PM_R <- cor(PM[-c(1,22)])
tgnEndo_R <- cor(tgnEndo[-c(1,22)])
```


### Corrplots
```{r corrplots}
#corrplots
pdf("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/ERcorrplot.pdf")
corrplot(ER_R, title = "Endoplasmic Reticulum", mar=c(0,0,2,0), order = "hclust", addrect = 4)
dev.off()

pdf("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/Golgicorrplot.pdf")
corrplot(golgi_R, title = "Golgi", mar=c(0,0,2,0), order = "hclust", addrect = 7)
dev.off()

pdf("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/Lysocorrplot.pdf")
corrplot(lyso_R, title = "Lysosomes", mar=c(0,0,2,0), order = "hclust", addrect = 7)
dev.off()

pdf("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/PMcorrplot.pdf")
corrplot(PM_R, title = "Plasma Membrane", mar=c(0,0,2,0), order = "hclust", addrect = 5)
dev.off()

pdf("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/TGNEndocorrplot.pdf")
corrplot(tgnEndo_R, title = "TGN:Endosomes", mar=c(0,0,2,0), order = "hclust", addrect = 4)
dev.off()

# mar=c(0,0,2,0)
#https://stackoverflow.com/questions/40509217/how-to-have-r-corrplot-title-position-correct

```


### MVN? There are no MVN outliers in the raw data
### None of the amino acids are univariate normal across all locations
### ER: only predictors that are univariate normal are: I, L, V, F, C, A, G, P
### Golgi: L, W, C, A, G, T, S
### Lysosomes: V
### PM: L, C, G, T, M
### TGN: A
```{r mvn}
data %>% select(I:Y) %>%
   ggpairs(title = "Pair plot of hydrophobic amino acids", lower = list(continuous = wrap("smooth", method = "lm", se = F, alpha = 0.5, color="steelblue")),
           diag = list(continuous = wrap("densityDiag", fill = "steelblue" )))
ggsave("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/pairs1.pdf")

data %>% select(C:M) %>%
   ggpairs(title = "Pair plot of neutral amino acids", lower = list(continuous = wrap("smooth", method = "lm", se = F,
                                          alpha = 0.5, color="steelblue")),
           diag = list(continuous = wrap("densityDiag", fill = "steelblue" )))
ggsave("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/pairs2.pdf")

data %>% select(H:R) %>%
   ggpairs(title = "Pair plot of charged amino acids", lower = list(continuous = wrap("smooth", method = "lm", se = F,
                                          alpha = 0.5, color="steelblue")),
           diag = list(continuous = wrap("densityDiag", fill = "steelblue" )))
ggsave("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/pairs3.pdf")

mvn(data=ER[-c(1,22)], mvnTest="mardia", multivariatePlot="qq")
mvn(data=golgi[-c(1,22)], mvnTest="mardia", multivariatePlot="qq")
mvn(data=lyso[-c(1,22)], mvnTest="mardia", multivariatePlot="qq")
mvn(data=PM[-c(1,22)], mvnTest="mardia", multivariatePlot="qq")
mvn(data=tgnEndo[-c(1,22)], mvnTest="mardia", multivariatePlot="qq")
```

### Log transformation
```{r log}
data[data == 0] <- 0.000001 # replacing 0s with a small number
log_data <- log(data[c(2:21)])
log_data <- cbind(data[c(1,22)], log_data)
log_data$Position <- as.integer(log_data$Position)
```

### MVN of log transformed data? Logit transformation achieved same regarding establishing MVN as log
```{r mvn log}
logER <- log_data %>% filter(location == "ER")
loggolgi <- log_data %>% filter(location == "Golgi")
logtgnEndo <- log_data %>% filter(location == "TGNEndo")
loglyso <- log_data %>% filter(location == "lysosomes")
logPM <- log_data %>% filter(location == "PM_total")

log_data %>% select(I:Y) %>%
   ggpairs(title = "Log transformed: hydrophobic amino acids", lower = list(continuous = wrap("smooth", method = "lm", se = F,
                                          alpha = 0.5, color="steelblue")),
           diag = list(continuous = wrap("densityDiag", fill = "steelblue" )))
ggsave("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/log_pairs1.pdf")

log_data %>% select(C:M) %>%
   ggpairs(title = "Log transformed: neutral amino acids", lower = list(continuous = wrap("smooth", method = "lm", se = F,
                                          alpha = 0.5, color="steelblue")),
           diag = list(continuous = wrap("densityDiag", fill = "steelblue" )))
ggsave("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/log_pairs2.pdf")

log_data %>% select(H:R) %>%
   ggpairs(title = "Log transformed: charged amino acids", lower = list(continuous = wrap("smooth", method = "lm", se = F,
                                          alpha = 0.5, color="steelblue")),
           diag = list(continuous = wrap("densityDiag", fill = "steelblue" )))
ggsave("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/log_pairs3.pdf")

mvn(data=logER[-c(1,2)], mvnTest="mardia", multivariateOutlierMethod = "quan", showOutliers = T, multivariatePlot="qq", showNewData = T)
golgiResult <- mvn(data=loggolgi[-c(1,2)], mvnTest="mardia", multivariateOutlierMethod = "quan", showOutliers = T,multivariatePlot="qq", showNewData = T)
mvn(data=loglyso[-c(1,2)], mvnTest="mardia", multivariateOutlierMethod = "quan", showOutliers = T, multivariatePlot="qq", showNewData = T)
mvn(data=logPM[-c(1,2)], mvnTest="mardia", multivariateOutlierMethod = "quan", showOutliers = T, multivariatePlot="qq", showNewData = T)
tgnResult <- mvn(data=logtgnEndo[-c(1,2)], mvnTest="mardia", multivariateOutlierMethod = "quan", showOutliers = T, multivariatePlot="qq", showNewData = T)

```

### Logit transformation
```{r logit}
data[data == 0] <- 0.000001 # replacing 0s with a small number
logit_data <- logit(data[c(2:21)])
logit_data <- cbind(data[c(1,22)], logit_data)
logit_data$Position <- as.integer(logit_data$Position)
```

### MVN of logit transformed data?
```{r mvn log}
logitER <- logit_data %>% filter(location == "ER")
logitgolgi <- logit_data %>% filter(location == "Golgi")
logittgnEndo <- logit_data %>% filter(location == "TGNEndo")
logitlyso <- logit_data %>% filter(location == "lysosomes")
logitPM <- logit_data %>% filter(location == "PM_total")

logit_data %>% select(I:Y) %>%
   ggpairs(lower = list(continuous = wrap("smooth", method = "loess", se = F,
                                          alpha = 0.5, color="steelblue")),
           diag = list(continuous = wrap("densityDiag", fill = "steelblue" )))

logit_data %>% select(C:M) %>%
   ggpairs(lower = list(continuous = wrap("smooth", method = "loess", se = F,
                                          alpha = 0.5, color="steelblue")),
           diag = list(continuous = wrap("densityDiag", fill = "steelblue" )))

logit_data %>% select(H:R) %>%
   ggpairs(lower = list(continuous = wrap("smooth", method = "loess", se = F,
                                          alpha = 0.5, color="steelblue")),
           diag = list(continuous = wrap("densityDiag", fill = "steelblue" )))

mvn(data=logitER[-c(1,2)], mvnTest="mardia", multivariateOutlierMethod = "quan", showOutliers = T, multivariatePlot="qq")
mvn(data=logitgolgi[-c(1,2)], mvnTest="mardia", multivariateOutlierMethod = "quan", showOutliers = T, multivariatePlot="qq")
mvn(data=logitlyso[-c(1,2)], mvnTest="mardia", multivariateOutlierMethod = "quan", showOutliers = T, multivariatePlot="qq")
mvn(data=logitPM[-c(1,2)], mvnTest="mardia", multivariateOutlierMethod = "quan", showOutliers = T, multivariatePlot="qq")
mvn(data=logittgnEndo[-c(1,2)], mvnTest="mardia", multivariateOutlierMethod = "quan", showOutliers = T, multivariatePlot="qq")
```

### no MVN outliers
```{r no outliers}
#indexing row names to exclude MVN outliers
Golgiidx <- as.integer(rownames(golgiResult$multivariateOutliers))
TGNidx <- as.integer(rownames(tgnResult$multivariateOutliers))

GolgiNoOut <- loggolgi[Golgiidx,]
TGNnoOut <- logtgnEndo[TGNidx,]

noOutData <- rbind(GolgiNoOut, TGNnoOut) %>% arrange(location, Position)
```


### Grouping MVN data without outliers for MANOVA
```{r mvn data}
mvnData <- rbind(logER, loglyso, logPM, noOutData) %>% arrange(location, Position)
fullset <- log_data %>% filter(location != "Golgi" | location != "TGNEndo")
boxM(cbind(I, L, V, F, W, Y, C, A, G, P, T, S, M, H, Q, N, E, D, K, R)~factor(location), data = fullset)
```

### Problem is that in the mvnData, not all positions are the same for each location because outliers were removed
### Which amino acids in the TM domain vary most by position per location?
Need anova(residue~position) for I:Y, C:M, and H:R
```{r manova}
hydros <- mvnData %>% select(I:Y)
neutral <- mvnData %>% select(C, M, Q, N, A, G)
polar <- mvnData %>% select(P:S) 
charged <- mvnData %>% select(H, E:R)

hydros <- cbind(mvnData[c(1,22)], hydros)
neutral <- cbind(mvnData[c(1,22)], neutral)
polar <- cbind(mvnData[c(1,22)], polar)
charged <- cbind(mvnData[c(1,22)], charged)

mvnER <- mvnData %>% filter(location == "ER")
mvnGolgi <- mvnData %>% filter(location == "Golgi")
mvnPM <- mvnData %>% filter(location == "PM_total")
mvnLyso <- mvnData %>% filter(location == "lysosomes")
mvnTGN <- mvnData %>% filter(location == "TGNEndo")

hydroMod<- manova(cbind(I, L, V, F, W, Y)~factor(location), data = hydros)#Fits the MANOVA model for meanvec of vars by location
summary(hydroMod, test="Wilks") 

neutMod <- manova(cbind(C, M, Q, N, A, G)~factor(location), data=neutral)#Fits the MANOVA model for meanvec of vars by location
summary(neutMod, test="Wilks") 

chargeMod <- manova(cbind(H, E, D, K, R)~factor(location), data=charged)#Fits the MANOVA model for meanvec of vars by location
summary(chargeMod, test="Wilks") 

polMod <- manova(cbind(P, T, S)~factor(location), data=polar)#Fits the MANOVA model for meanvec of vars by location
summary(polMod, test="Wilks") 

hydroModP<- manova(cbind(I, L, V, F, W, Y)~Position, data = hydros)#Fits the MANOVA model for meanvec of vars by location
summary(hydroModP, test="Wilks") 

neutModP <- manova(cbind(C, M, Q, N, A, G)~Position, data=neutral)#Fits the MANOVA model for meanvec of vars by location
summary(neutModP, test="Wilks") 

chargeModP <- manova(cbind(H, E, D, K, R)~Position, data=charged)#Fits the MANOVA model for meanvec of vars by location
summary(chargeModP, test="Wilks") 

polModP <- manova(cbind(P, T, S)~Position, data=polar)#Fits the MANOVA model for meanvec of vars by location
summary(polModP, test="Wilks") 
```

### Decomposing MANOVA - All amino acids differ by location
```{r finding diffs}
summary.aov(hydroMod)
summary.aov(neutMod)
summary.aov(polMod)
summary.aov(chargeMod)

summary.aov(hydroModP)
summary.aov(neutModP)
summary.aov(polMod)
summary.aov(chargeModP)

boxM(cbind(I, L, V, F, W, Y)~factor(location), data = hydros)
boxM(cbind(C, M, Q, N, A, G)~factor(location), data=neutral)
boxM(cbind(H, E, D, K, R)~factor(location), data=charged)
boxM(cbind(P, T, S)~factor(location), data=polar)

```

### Pairwise t tests
```{r t}
pairwise.t.test(mvnData$I, mvnData$location, p.adjust.method = "bonferroni")
pairwise.t.test(data$I, data$location, p.adjust.method = "bonferroni")
pairwise.t.test(log_data$I, log_data$location, p.adjust.method = "bonferroni")
```


### MANOVA with log transformed data with outliers present
```{r raw log_data}
hydros <- log_data %>% select(I:Y)
neutral <- log_data %>% select(C, M, Q, N, A, G)
polar <- log_data %>% select(P:S) 
charged <- log_data %>% select(H, E:R)

hydros <- cbind(log_data[c(1,2)], hydros)
neutral <- cbind(log_data[c(1,2)], neutral)
polar <- cbind(log_data[c(1,2)], polar)
charged <- cbind(log_data[c(1,2)], charged)

logER <- log_data %>% filter(location == "ER")
logGolgi <- log_data %>% filter(location == "Golgi")
logPM <- log_data %>% filter(location == "PM_total")
logLyso <- log_data %>% filter(location == "lysosomes")
logTGN <- log_data %>% filter(location == "TGNEndo")

hydroMod<- manova(cbind(I, L, V, F, W, Y)~factor(location), data = hydros)#Fits the MANOVA model for meanvec of vars by location
summary(hydroMod, test="Wilks") 

neutMod <- manova(cbind(C, M, Q, N, A, G)~factor(location), data=neutral)#Fits the MANOVA model for meanvec of vars by location
summary(neutMod, test="Wilks") 

chargeMod <- manova(cbind(H, E, D, K, R)~factor(location), data=charged)#Fits the MANOVA model for meanvec of vars by location
summary(chargeMod, test="Wilks") 

polMod <- manova(cbind(P, T, S)~factor(location), data=polar)#Fits the MANOVA model for meanvec of vars by location
summary(polMod, test="Wilks") 

hydroModP<- manova(cbind(I, L, V, F, W, Y)~Position, data = hydros)#Fits the MANOVA model for meanvec of vars by location
summary(hydroModP, test="Wilks") 

neutModP <- manova(cbind(C, M, Q, N, A, G)~Position, data=neutral)#Fits the MANOVA model for meanvec of vars by location
summary(neutModP, test="Wilks") 

chargeModP <- manova(cbind(H, E, D, K, R)~Position, data=charged)#Fits the MANOVA model for meanvec of vars by location
summary(chargeModP, test="Wilks") 

polModP <- manova(cbind(P, T, S)~Position, data=polar)#Fits the MANOVA model for meanvec of vars by location
summary(polModP, test="Wilks") 
```

### Decomposing raw log MANOVA
```{r ind ANOVA log data}
summary.aov(hydroMod)
summary.aov(neutMod)
summary.aov(polMod)
summary.aov(chargeMod)

summary.aov(hydroModP)
summary.aov(neutModP)
summary.aov(polMod)
summary.aov(chargeModP)
```

### ER PCA
```{r ER_pca}
ER_PCA <- prcomp(ER[-c(1,22)], scale. = T)
summary(ER_PCA)

pER <- ncol(ER[-c(1,22)])
ERPCdat <- data.frame(var_S = diag(ER_S),
           PC1 = ER_PCA$rotation[,1],
           PC2 = ER_PCA$rotation[,2],
           avg_cor = colMeans(abs(ER_R))-1/pER) %>% # subtracting 1/p removes the correlation of the variable with itself
  arrange(desc(PC1)) #PC1: Hydrophobics at one end, charged at the other
                     #PC1: Looks like it's based on decreasing then increasing avg.                           correlation
                     #PC2: not sure what its basis is
ERPCdat <-rownames_to_column(ERPCdat, var = "sidechain")

ggplot(data = ERPCdat, mapping = aes(x=PC2, y=PC1, color = sidechain, label  = sidechain)) +
  scale_color_manual(values = palette) +
  geom_point(size = 2) +
  geom_text(hjust = 0, nudge_x = 0.01, nudge_y = 0.01) +
  labs(title = "Plot of ER Principal Components") +
  theme_minimal(base_size = 18) +
  theme(legend.position = "none")
#ggsave("/Users/leekapp/Documents/Portfolio/STAT223Final/images/ERPCplot.pdf", width = 6, height = 6, units = "in")
```

### Golgi PCA
```{r golgi_PCA}
golgi_PCA <- prcomp(golgi[-c(1,22)], scale. = T)
summary(golgi_PCA)

pG <- ncol(golgi[-c(1,22)])
golgi_dat <- data.frame(var_S = diag(golgi_S),
           PC1 = golgi_PCA$rotation[,1],
           PC2 = golgi_PCA$rotation[,2],
           avg_cor = colMeans(abs(golgi_R))-1/pG) %>% # subtracting 1/p removes the correlation of the variable with itself
  arrange(desc(PC1)) #PC1: Hydrophobics at one end, charged at the other
                     #PC1: Looks like it's based on decreasing then increasing avg.                           correlation
                     #PC2: not sure what its basis is
golgi_dat <-rownames_to_column(golgi_dat, var = "sidechain")

ggplot(data = golgi_dat, mapping = aes(x=PC2, y=PC1, color = sidechain, label = sidechain)) +
  scale_color_manual(values = palette) +
  geom_point(size = 3) +
  geom_text(hjust = 0, nudge_x = 0.0025, nudge_y = 0.025) +
  labs(title = "Plot of Golgi Principal Components") +
  theme_minimal(base_size = 18)  +
  theme(legend.position = "none")
#ggsave("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/GolgiPCplot.pdf", width = 6, height = 6, units = "in")
```

### PM_PCA
```{r PM_PCA}
PM_PCA <- prcomp(PM[-c(1,22)], scale. = T)
summary(PM_PCA)

pPM <- ncol(PM[-c(1,22)])
PM_dat <- data.frame(var_S = diag(PM_S),
           PC1 = PM_PCA$rotation[,1],
           PC2 = PM_PCA$rotation[,2],
           avg_cor = colMeans(abs(PM_R))-1/pPM) %>% # subtracting 1/p removes the correlation of the variable with itself
  arrange(desc(PC1)) #PC1: Hydrophobics at one end, charged at the other
                     #PC1: Looks like it's based on decreasing then increasing avg.                           correlation
                     #PC2: not sure what its basis is
PM_dat <-rownames_to_column(PM_dat, var = "sidechain")

ggplot(data = PM_dat, mapping = aes(x=PC2, y=PC1, color = sidechain, label = sidechain)) +
  scale_color_manual(values = palette) +
  geom_point(size = 3) +
  geom_text(hjust = 0, nudge_x = 0.01, nudge_y = 0.01) +
  labs(title = "Plot of Plasma Membrane Principal Components") +
  theme_minimal(base_size = 18) +
  theme(legend.position = "none")
#ggsave("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/PMPCplot.pdf", width = 8, height = 6, units = "in")
```

### Lysosome_PCA
```{r lyso_PCA}
lyso_PCA <- prcomp(lyso[-c(1,22)], scale. = T)
summary(lyso_PCA)

pL <- ncol(lyso[-c(1,22)])
Lys_dat <- data.frame(var_S = diag(lyso_S),
           PC1 = lyso_PCA$rotation[,1],
           PC2 = lyso_PCA$rotation[,2],
           avg_cor = colMeans(abs(lyso_R))-1/pL) %>% # subtracting 1/p removes the correlation of the variable with itself
  arrange(desc(PC1)) #PC1: Hydrophobics at one end, charged at the other
                     #PC1: Looks like it's based on decreasing then increasing avg.                           correlation
                     #PC2: not sure what its basis is

Lys_dat <-rownames_to_column(Lys_dat, var = "sidechain")

ggplot(data = Lys_dat, mapping = aes(x=PC2, y=PC1, color = sidechain, label = sidechain)) +
  scale_color_manual(values = palette) +
  geom_point(size = 3) +
  geom_text(hjust = 0, nudge_x = 0.01, nudge_y = 0.01) +
  labs(title = "Plot of Lysosome Principal Components") +
  theme_minimal(base_size = 18) +
  theme(legend.position = "none")
#ggsave("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/LysPCplot.pdf", width = 8, height = 6, units = "in")

```

### TGN_Endo_PCA
```{r  tgnEndo_PCA}
tgnEndo_PCA <- prcomp(tgnEndo[-c(1,22)], scale. = T)
summary(tgnEndo_PCA)

pT <- ncol(tgnEndo[-c(1,22)])
TGN_dat <- data.frame(var_S = diag(tgnEndo_S),
           PC1 = tgnEndo_PCA$rotation[,1],
           PC2 = tgnEndo_PCA$rotation[,2],
           avg_cor = colMeans(abs(tgnEndo_R))-1/pT) %>% # subtracting 1/p removes the correlation of the variable with itself
  arrange(desc(PC1)) #PC1: Hydrophobics at one end, charged at the other but in opposite                      order relative to other compartments!
                     #PC1: Looks like it's based on decreasing then increasing avg.                           correlation
                     #PC2: not sure what its basis is
TGN_dat <-rownames_to_column(TGN_dat, var = "sidechain")

ggplot(data = TGN_dat, mapping = aes(x=PC2, y=PC1, color = sidechain, label = sidechain)) +
  scale_color_manual(values = palette) +
  geom_point(size = 3) +
  geom_text(hjust = 0, nudge_x = 0.01, nudge_y = 0.01) +
  labs(title = "Plot of TGN:Endosome Principal Components") +
  theme_minimal(base_size = 18) +
  theme(legend.position = "none")
#ggsave("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/TGNPCplot.pdf", width = 8, height = 6, units = "in")
```

### Screeplots
```{r screeplots}
pdf("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/ERscree.pdf")
screeplot(ER_PCA,  npcs=pER, type="lines", pch=19, col=4, main = "ER PCA")
dev.off()
pdf("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/Golgiscree.pdf")
screeplot(golgi_PCA,  npcs=pG, type="lines", pch=19, col=4, main = "Golgi PCA")
dev.off()
pdf("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/PMscree.pdf")
screeplot(PM_PCA,  npcs=pPM, type="lines", pch=19, col=4, main = "Plasma Membrane PCA")
dev.off()
pdf("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/Lysoscree.pdf")
screeplot(lyso_PCA,  npcs=pL, type="lines", pch=19, col=4, main = "Lysosomes PCA")
dev.off()
pdf("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/TGNEndoscree.pdf")
screeplot(tgnEndo_PCA,  npcs=pT, type="lines", pch=19, col=4, "TGN:Endosomes PCA")
dev.off()
```

### PCA Tables - does PC2 sort based on charge, size, linear vs ring, or some other chemical property?
```{r PCATables}
colnames(ERPCdat) <- c("Sidechain", "Variance", "PC1", "PC2", "Average Corr.")
colnames(golgi_dat) <- c("Sidechain", "Variance", "PC1", "PC2", "Average Corr.")
colnames(Lys_dat) <- c("Sidechain", "Variance", "PC1", "PC2", "Average Corr.")
colnames(PM_dat) <- c("Sidechain", "Variance", "PC1", "PC2", "Average Corr.")
colnames(TGN_dat) <- c("Sidechain", "Variance", "PC1", "PC2", "Average Corr.")

ERPCdat %>% 
  kbl(caption = "Table 1. Examination of ER Principal Components") %>% 
  kable_classic(full_width = F, html_font = "Courier", font_size = 12) %>% 
  column_spec(1, bold = T) %>% 
  row_spec(0, bold = T) %>% 
  save_kable("/Users/leekapp/Documents/Portfolio/STAT223Final/images/ERPCtable.pdf")

golgi_dat %>% 
  kbl(caption = "Table 2. Examination of Golgi Principal Components") %>% 
  kable_classic(full_width = F, html_font = "Courier", font_size = 12) %>% 
  column_spec(1, bold = T) %>% 
  row_spec(0, bold = T) %>% 
  save_kable("/Users/leekapp/Documents/Portfolio/STAT223Final/images/GolgiPCtable.pdf")

Lys_dat %>% 
  kbl(caption = "Table 3. Examination of Lysosome Principal Components") %>% 
  kable_classic(full_width = F, html_font = "Courier", font_size = 12) %>% 
  column_spec(1, bold = T) %>% 
  row_spec(0, bold = T) %>% 
  save_kable("/Users/leekapp/Documents/Portfolio/STAT223Final/images/LysPCtable.pdf")

PM_dat %>% 
  kbl(caption = "Table 4. Examination of Plasma Membrane Principal Components") %>% 
  kable_classic(full_width = F, html_font = "Courier", font_size = 12) %>% 
  column_spec(1, bold = T) %>% 
  row_spec(0, bold = T) %>% 
  save_kable("/Users/leekapp/Documents/Portfolio/STAT223Final/images/PMPCtable.pdf")

TGN_dat %>% 
  kbl(caption = "Table 5. Examination of TGN:Endosome Principal Components") %>% 
  kable_classic(full_width = F, html_font = "Courier", font_size = 12) %>% 
  column_spec(1, bold = T) %>% 
  row_spec(0, bold = T) %>% 
  save_kable("/Users/leekapp/Documents/Portfolio/STAT223Final/images/TGNPCtable.pdf")

```


### Factor Analysis - choosing m
```{r chooseM}
fa.parallel(ER[-c(1,22)], fm = "pa", fa="fa")
fa.parallel(golgi[-c(1,22)], fm = "pa", fa="fa")
fa.parallel(lyso[-c(1,22)], fm = "pa", fa="fa")
fa.parallel(tgnEndo[-c(1,22)], fm = "pa", fa="fa")
fa.parallel(PM[-c(1,22)], fm = "pa", fa="fa")
```


### PC Method - ER - Looks like 6 factors are needed based on residuals
```{r FA_ER}
(ER_pcfa6 <- fa(r = cor(ER[-c(1,22)]),
                  nfactors = 6,
                  fm="pa",
                  rotate = "none",
                  n.obs = n))
summary(ER_pcfa6)
round(ER_pcfa6$residual, 3)
res_mat6 <- ER_pcfa6$residual-diag(diag(ER_pcfa6$residual))
max(abs(res_mat6))

corrplot(ER_pcfa6$residual, type = "full",
         method = "pie")

(ER_pcfa5 <- fa(r = cor(ER[-c(1,22)]), 
                  nfactors = 5, 
                  fm="pa",
                  rotate = "none",
                  n.obs = n))
summary(ER_pcfa5)
round(ER_pcfa5$residual, 3)
res_mat5 <- ER_pcfa5$residual-diag(diag(ER_pcfa5$residual))
max(abs(res_mat5))

corrplot(ER_pcfa5$residual, type = "full", 
         method = "pie")
```


### PC Method - golgi - Looks like 5 factors are needed based on residuals
```{r golgi_FA}
(golgi_pcfa5 <- fa(r = cor(golgi[-c(1,22)]),
                  nfactors = 5,
                  fm="pa",
                  rotate = "none",
                  n.obs = n))
summary(golgi_pcfa5)
round(golgi_pcfa5$residual, 3)
res_mat5 <- golgi_pcfa5$residual-diag(diag(golgi_pcfa5$residual))
max(abs(res_mat5))

corrplot(golgi_pcfa5$residual, type = "full",
         method = "pie")

(golgi_pcfa4 <- fa(r = cor(golgi[-c(1,22)]),
                  nfactors = 4,
                  fm="pa",
                  rotate = "none",
                  n.obs = n))
summary(golgi_pcfa4)
round(golgi_pcfa4$residual, 3)
res_mat4 <- golgi_pcfa4$residual-diag(diag(golgi_pcfa4$residual))
max(abs(res_mat4))

corrplot(golgi_pcfa4$residual, type = "full",
         method = "pie")
```


### PC Method Lysosomes - looks like 6 factors are needed based on residuals 
```{r lyso_FA}
(lyso_pcfa5 <- fa(r = cor(lyso[-c(1,22)]),
                  nfactors = 5,
                  fm="pa",
                  rotate = "none",
                  n.obs = n))
summary(lyso_pcfa5)
round(lyso_pcfa5$residual, 3)
res_mat5 <- lyso_pcfa5$residual-diag(diag(lyso_pcfa5$residual))
max(abs(res_mat5))

corrplot(lyso_pcfa5$residual, type = "full",
         method = "pie")

(lyso_pcfa6 <- fa(r = cor(lyso[-c(1,22)]),
                  nfactors = 6,
                  fm="pa",
                  rotate = "none",
                  n.obs = n))
summary(lyso_pcfa6)
round(lyso_pcfa6$residual, 3)
res_mat6 <- lyso_pcfa6$residual-diag(diag(lyso_pcfa6$residual))
max(abs(res_mat6))

corrplot(lyso_pcfa6$residual, type = "full",
         method = "pie")
```


### PC Method - PM - maybe 5, definitely 4 factors needed
```{r PM_FA}
(PM_pcfa4 <- fa(r = cor(PM[-c(1,22)]),
                  nfactors = 4,
                  fm="pa",
                  rotate = "none",
                  n.obs = n))
summary(PM_pcfa4)
round(PM_pcfa4$residual, 3)
res_mat5 <- PM_pcfa4$residual-diag(diag(PM_pcfa4$residual))
max(abs(res_mat4))

corrplot(PM_pcfa4$residual, type = "full",
         method = "pie")

(PM_pcfa5 <- fa(r = cor(PM[-c(1,22)]),
                  nfactors = 5,
                  fm="pa",
                  rotate = "none",
                  n.obs = n))
summary(PM_pcfa5)
round(PM_pcfa5$residual, 3)
res_mat5 <- PM_pcfa5$residual-diag(diag(PM_pcfa5$residual))
max(abs(res_mat5))

corrplot(PM_pcfa5$residual, type = "full",
         method = "pie")
```


### PC Method - TGN:Endosomes - 5 to 7 factors are required based on residuals
```{r PM_FA}
(tgn_pcfa7 <- fa(r = cor(tgnEndo[-c(1,22)]),
                  nfactors = 7,
                  fm="pa",
                  rotate = "none",
                  n.obs = n))
summary(tgn_pcfa7)
round(tgn_pcfa7$residual, 3)
res_mat7 <- tgn_pcfa7$residual-diag(diag(tgn_pcfa7$residual))
max(abs(res_mat7))

corrplot(tgn_pcfa7$residual, type = "full",
         method = "pie")

(tgn_pcfa5 <- fa(r = cor(tgnEndo[-c(1,22)]),
                  nfactors = 5,
                  fm="pa",
                  rotate = "none",
                  n.obs = n))
summary(tgn_pcfa5)
round(tgn_pcfa5$residual, 3)
res_mat5 <- tgn_pcfa5$residual-diag(diag(tgn_pcfa5$residual))
max(abs(res_mat5))

corrplot(tgn_pcfa5$residual, type = "full",
         method = "pie")
```


### Visualizing FA
```{r}
theme_set(theme_classic()+
             theme(axis.title = element_text(size = 16),
                   axis.text = element_text(size = 16),
                   plot.title = element_text(size = 18, hjust = 0.5),
                   plot.subtitle = element_text(size = 14, hjust = 0.5),
                   plot.caption = element_text(size = 12)))


plot_theme <- theme(
   axis.title = element_text(size = 16),
   axis.text = element_text(size = 16),
   plot.title = element_text(size = 18, hjust = 0.5),
   plot.subtitle = element_text(size = 14, hjust = 0.5),
   plot.caption = element_text(size = 12),
   legend.position = "none")

# Convert the loadings to a dataframe
data.frame(ER_pcfa6$loadings[,1:3]) %>%
   
   # Changing the variables from a row name to a column named variable
   rownames_to_column(var="Variable") %>%
   
   # Call on ggplot() and specify x = 1st and y = 2nd factor
   ggplot(aes(x = PA1, y = PA2, color = Variable)) + 
   geom_point(aes(size=PA3)) + 
   scale_color_manual(values = palette) +
   geom_text_repel(aes(label = Variable)) +
   labs(x = "Factor 1", y = "Factor 2",
        title = "Factor Analysis for ER Data") +
   theme_bw() + plot_theme
ggsave("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/ERFAplot.pdf", width = 6, height = 6, units = "in")

data.frame(golgi_pcfa5$loadings[,1:3]) %>%
   
   # Changing the variables from a row name to a column named variable
   rownames_to_column(var="Variable") %>%
   
   # Call on ggplot() and specify x = 1st and y = 2nd factor
   ggplot(aes(x = PA1, y = PA2, color = Variable)) + 
   geom_point(aes(size=PA3)) + 
   scale_color_manual(values = palette) +
   geom_text_repel(aes(label = Variable)) +
   labs(x = "Factor 1", y = "Factor 2",
        title = "Factor Analysis for Golgi Data") +
   theme_bw() + plot_theme
ggsave("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/golgiFAplot.pdf", width = 6, height = 6, units = "in")

data.frame(lyso_pcfa6$loadings[,1:3]) %>%
   
   # Changing the variables from a row name to a column named variable
   rownames_to_column(var="Variable") %>%
   
   # Call on ggplot() and specify x = 1st and y = 2nd factor
   ggplot(aes(x = PA1, y = PA2, color = Variable)) + 
   geom_point(aes(size=PA3)) + 
   scale_color_manual(values = palette) +
   geom_text_repel(aes(label = Variable)) +
   labs(x = "Factor 1", y = "Factor 2",
        title = "Factor Analysis for Lysosomes Data") +
   theme_bw() + plot_theme
ggsave("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/lysoFAplot.pdf", width = 6, height = 6, units = "in")

data.frame(PM_pcfa5$loadings[,1:3]) %>%
   
   # Changing the variables from a row name to a column named variable
   rownames_to_column(var="Variable") %>%
   
   # Call on ggplot() and specify x = 1st and y = 2nd factor
   ggplot(aes(x = PA1, y = PA2, color = Variable)) + 
   geom_point(aes(size=PA3)) + 
   scale_color_manual(values = palette) +
   geom_text_repel(aes(label = Variable)) +
   labs(x = "Factor 1", y = "Factor 2",
        title = "Factor Analysis for Plasma Membrane Data") +
   theme_bw() + plot_theme
ggsave("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/PMFAplot.pdf", width = 6, height = 6, units = "in")

data.frame(tgn_pcfa5$loadings[,1:3]) %>%
   
   # Changing the variables from a row name to a column named variable
   rownames_to_column(var="Variable") %>%
   
   # Call on ggplot() and specify x = 1st and y = 2nd factor
   ggplot(aes(x = PA1, y = PA2, color = Variable)) + 
   geom_point(aes(size=PA3)) + 
   scale_color_manual(values = palette) +
   geom_text_repel(aes(label = Variable)) +
   labs(x = "Factor 1", y = "Factor 2",
        title = "Factor Analysis for TGN:Endosomes Data") +
   theme_bw() + plot_theme
ggsave("/Users/leekapp/Desktop/STAT223/Assignments/STAT223FinalProject/tgnFAplot.pdf", width = 6, height = 6, units = "in")
```

### K-means
```{r kmeans}
### K-means clustering
AA_stan = data %>% select(-Position, -location) %>% mutate_all(scale)

fviz_pca_ind(prcomp(AA_stan), geom="text")
fviz_nbclust(AA_stan, FUNcluster = kmeans, method="wss") 
fviz_nbclust(AA_stan, FUNcluster = kmeans, method="silhouette")
# Use 4 groups

(AA_km <- kmeans(x=AA_stan, centers= 5, nstart=20))
fviz_cluster(AA_km, AA_stan, repel=T)
cluster <- AA_km$cluster

clust_data <- cbind(data, cluster)
AA_km$centers
```

